

---
title: "Advanced Statistical Analysis of Anxiety and Depression Trends During the
  COVID-19 Pandemic"
author: "Ronak Fathi"
date: "`r Sys.Date()`"
output:
  pdf_document: default
  html_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE)
library(tidyverse)
library(lubridate)
library(scales)
library(readr)
library(mgcv)        # GAM
library(lme4)        # Mixed models
library(nlme)        # Mixed models (alternative)
library(car)         # MANOVA
library(cluster)     # Clustering
library(factoextra)  # PCA and clustering viz
library(psych)       # PCA support
library(broom)
library(ggpubr)
```

# Abstract

This study applies advanced statistical techniques to examine trends in self-reported anxiety and depression symptoms during the COVID-19 pandemic using the CDC Household Pulse Survey. We use generalized linear and additive models, mixed effects models, time series methods, multivariate techniques like PCA and MANOVA, and clustering to understand how mental health patterns evolved over time and across demographic subgroups.

# Introduction

The COVID-19 pandemic has profoundly impacted public mental health, leading to increased levels of anxiety and depression. To assess these trends, the U.S. Census Bureau and National Center for Health Statistics launched the Household Pulse Survey. This analysis uses advanced statistical modeling to explore temporal dynamics and subgroup differences in mental health outcomes based on this large-scale repeated cross-sectional dataset.

# Data Description
```{r load-data}
setwd("C:/Users/0&1/OneDrive/Documents/R-projects/Covid")
df_raw <- read.csv("Indicators_of_Anxiety_or_Depression.csv")
```

```{r clean}
df <- df_raw %>%
  rename_with(~ gsub(" ", "_", .x)) %>%
  rename(
    indicator = Indicator,
    group = Group,
    subgroup = Subgroup,
    value = Value,
    low_ci = LowCI,
    high_ci = HighCI,
    week = Phase,
    start_date = TimePeriodStartDate,
    end_date = TimePeriodEndDate
  ) %>%
  mutate(
    start_date = mdy(start_date),  # fix here
    end_date = mdy(end_date),      # and here
    time_numeric = as.numeric(start_date)
  ) %>%
  filter(indicator == "Symptoms of Anxiety Disorder or Depressive Disorder") %>%
  drop_na(value)
```


# Exploratory Analysis

```{r eda}
ggplot(df %>% filter(group == "National Estimate"), aes(x = start_date, y = value)) +
  geom_line(color = "steelblue") +
  labs(title = "National Trend in Reported Anxiety/Depression", x = "Date", y = "Percentage") +
  theme_minimal()
```
\newpage

# Generalized Linear Model (GLM)

```{r glm}
glm_model <- glm(value ~ time_numeric, data = df %>% filter(group == "National Estimate"), family = gaussian())
summary(glm_model)
```

## Interpretation:
- The coefficient for `time_numeric` shows the average change per day.
- A significant positive slope suggests increasing symptoms over time.

# Generalized Additive Model (GAM)

```{r gam}
gam_model <- gam(value ~ s(time_numeric), data = df %>% filter(group == "National Estimate"))
summary(gam_model)
plot(gam_model, se = TRUE, col = "blue", main = "GAM: Nonlinear Trend")
```

## Interpretation:
- `s(time_numeric)` models smooth nonlinear trends.
- EDF > 1 confirms nonlinearity; p-value shows the trend is significant.
\newpage

# Linear Mixed Effects Model (LMM)

```{r lmm}
lmm_data <- df %>% filter(group %in% c("By Age", "By Race/Hispanic ethnicity"))
lmm <- lmer(value ~ time_numeric + (1|subgroup), data = lmm_data)
summary(lmm)
```

## Interpretation:
- Fixed effect = average change over time.
- Random intercepts allow baseline variation across subgroups.
\newpage

# Time Series Analysis (LOESS)

```{r ts-loess}
ts_data <- df %>%
  filter(group == "National Estimate") %>%
  group_by(start_date) %>%
  summarise(mean_value = mean(value))

ggplot(ts_data, aes(x = start_date, y = mean_value)) +
  geom_point() +
  geom_smooth(method = "loess", span = 0.3, se = FALSE, color = "darkred") +
  labs(title = "LOESS Smoothed Trend", y = "Avg % Reporting Symptoms")
```

## Interpretation:
- LOESS identifies short-term peaks and dips over time.
\newpage

# MANOVA

```{r manova}
# Filter for Age and Race groups
manova_df <- df %>%
  filter(group %in% c("By Age", "By Race/Hispanic ethnicity"),
         !is.na(value)) %>%
  select(time_numeric, group, subgroup, value)

# Pivot to wide format: subgroup becomes columns
manova_wide <- manova_df %>%
  unite(varname, group, subgroup, sep = "_") %>%
  pivot_wider(names_from = varname, values_from = value) %>%
  drop_na()

# View shape
nrow(manova_wide)  # should be > 0
head(manova_wide)

# Run MANOVA: test effect of time on the set of mental health outcomes by age/race groups
response_vars <- manova_wide %>% select(-time_numeric)
fit <- manova(as.matrix(response_vars) ~ time_numeric, data = manova_wide)

# Summary with Wilks' lambda
summary(fit, test = "Wilks")


```

## Interpretation:
- Wilks' Lambda tests joint differences across time.
- Significant result indicates meaningful group-level changes.
\newpage

# Principal Component Analysis (PCA)

```{r pca}
# Prepare data for PCA
pca_data <- manova_wide %>%
  select(-time_numeric)  # only the mental health response variables

# Remove columns with zero variance (constant variables)
pca_data <- pca_data %>% select(where(~ sd(., na.rm = TRUE) > 0))

# Run PCA with scaling
pca_result <- prcomp(pca_data, scale. = TRUE)

# View summary of explained variance
summary(pca_result)

# Scree plot
library(ggplot2)
scree_df <- data.frame(PC = paste0("PC", 1:length(pca_result$sdev)),
                       Variance = pca_result$sdev^2 / sum(pca_result$sdev^2))
ggplot(scree_df, aes(x = PC, y = Variance)) +
  geom_col(fill = "steelblue") +
  theme_minimal() +
  labs(title = "Scree Plot", y = "Proportion of Variance Explained")


# Create a data frame of the first few principal components for clustering
pca_df <- as.data.frame(pca_result$x[, 1:3])  # Use top 3 PCs
```
## Interpretation:
- PCA reveals latent patterns in subgroup responses.
- PC1 usually captures general trend; others may reflect specific group contrasts.
\newpage

# Clustering Analysis (K-Means)

```{r clustering}
clust_data <- scale(na.omit(pca_df))
k2 <- kmeans(clust_data, centers = 2, nstart = 25)
fviz_cluster(k2, data = clust_data, main = "K-Means Clustering")
```

## Interpretation:
- Clustering groups similar subgroup response patterns.
- Useful for targeting interventions to specific demographic clusters.
\newpage

# Model Diagnostics

```{r diagnostics}
par(mfrow = c(2, 2))
plot(glm_model)
#plot(gam_model, residuals = TRUE, pch = 20)
```

## Interpretation:
- Check assumptions: normality, homoscedasticity, outliers.
- GAM plots also show if smooths are well-fit.
\newpage

# Conclusion

The analysis revealed complex, nonlinear trends in self-reported mental health outcomes during the pandemic. Generalized Additive Models captured smooth variation, while Mixed Effects Models showed heterogeneity across subgroups. MANOVA confirmed significant temporal changes by age and race. PCA and clustering exposed latent patterns and helped classify at-risk groups. These findings underscore the importance of advanced analytics for understanding large public health datasets.

# References

- CDC Household Pulse Survey: [https://www.cdc.gov/nchs/covid19/pulse/mental-health.htm](https://www.cdc.gov/nchs/covid19/pulse/mental-health.htm)
- Wood, S. N. (2017). *Generalized Additive Models: An Introduction with R*.
- Pinheiro & Bates (2000). *Mixed-Effects Models in S and S-PLUS*.
- Jolliffe, I. T., & Cadima, J. (2016). *Principal component analysis: a review and recent developments*.
